---
title: "Case study of using survextrap: cetuximab for head and neck cancer"
author: "Christopher Jackson <chris.jackson@mrc-bsu.cam.ac.uk>"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_document:
    toc: true
    keep_md: true
    self_contained: true
    embed_resources: true
    toc_float: true
    theme: simplex
    code_folding: show
vignette: >
  %\VignetteIndexEntry{Case study of using survextrap: cetuximab for head and neck cancer}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
pkgdown:
  as_is: true
---

```{r 'setup', echo = FALSE, cache = FALSE}
knitr::opts_chunk$set(
  ## Get rid of those -1 at the end of the figure filenames
  ## Assumes there is only one figure per chunk
  fig.process = function(filename) {
    new_filename <- stringr::str_remove(string = filename,
                                        pattern = "-1")
    fs::file_move(path = filename, new_path = new_filename)
    ifelse(fs::file_exists(new_filename), new_filename, filename)
  },
  dev = c('png', 'pdf'),
  fig.align = 'center', 
  fig.height = 5, 
  fig.width = 8.5, 
  pdf.options(encoding = "ISOLatin9.enc")
)
```

This article is an in-depth demonstration of how `survextrap` could be used in a realistic application to a health technology evaluation.  A range of data sources and analysis choices are explored, and publication-style plots and tables are produced.

A more formal account of this analysis is given in the preprint [paper](https://arxiv.org/abs/2306.03957) about the package and method.   This article also serves the purpose of providing the code that was used in the paper, making the analysis reproducible and explaining how the code is used.

A more casual, quick tour of the package's features is given in the `survextrap` package vignette called [examples.html](examples.html), and the methods are explained in detail in [methods.html](methods.html).


# Simple data summary 

The data are from an analysis of survival in head and neck cancer, previously conducted by [Guyot et al.](https://journals.sagepub.com/doi/pdf/10.1177/0272989X16670604).  The `survextrap` package provides the data as `cetux`, `cetux_seer` and `cetux_bh`.   `cetux` is data from a clinical trial of cetuximab and radiotherapy, compared to a control group who only received radiotherapy.   `cetux_seer` contain matched data from a USA cancer registry, and `cetux_bh` describes comparable survival for the general population of the USA.   See `help("cetux")` and [Guyot et al.](https://journals.sagepub.com/doi/pdf/10.1177/0272989X16670604) for more information. 

We first load the required packages, plot a Kaplan-Meier curve of the individual trial data, and create subsets of the data by treatment group, for use in some of the later analyses.

```{r packages_data,warning=FALSE,message=FALSE}
library(survextrap)
library(ggplot2)
library(dplyr)
library(viridis) # for colour palettes
survminer::ggsurvplot(survfit(Surv(years, d) ~ treat, data=cetux)) + xlab("Years")
control <- cetux[cetux$treat=="Control",]
cetuximab <- cetux[cetux$treat=="Cetuximab",]
```

# Spline specification

First we set up an M-spline, specifying only that we expect a maximum of 6 spline coefficients to be sufficient to describe hazard variations, and we want to allow hazards to change up to 20 years.  The resulting default spline knots are saved in the `mspline` object here.  The default knots are based on quantiles of the event times in the data (which is why we supply the data in this command).

```{r}
mspline <- mspline_spec(Surv(years, d) ~ 1, data=cetux, df=6, add_knots=20)
```

# Choice of priors

For transparency here, all priors are specified explicitly, rather than relying on the package defaults.   The parameters of spline models hard to interpret, but we can determine suitable priors by translating them to quantities that are easier to interpret.  This will sometimes require simulation.    Note also that knowledge of the spline knots is necessary to do some of these calculations, which is why we set up the `mspline` object in advance of setting up the priors.   See the linked function documentation for more details about how these functions work. 

#### Hazard scale parameter $\eta$.

Since patients in the trial have a median age of 57 (range 34 to 83), the prior for $\eta$ is calibrated to imply a prior mean survival of 25 years after diagnosis, but with a variance chosen so that mean survival could be as high as 100 years after diagnosis.     The function `p_meansurv()` translates a prior median and upper 95\% credible limit for mean survival to an appropriate normal distribution for the log hazard scale $\eta$.  Note that this distribution represents uncertainty about the population mean, not individual variability (which should be broader).

Conversely, the `prior_haz_const()` function translates a normal prior for $\eta$ to the corresponding beliefs about survival.  We use this to check that the lower limit of the distribution derived by `p_meansurv()` is sensible. 

```{r prior_hscale}
prior_hscale <- p_meansurv(median=25, upper=100, mspline=mspline)
prior_haz_const(mspline, prior_hscale = prior_hscale)
```

#### Hazard variability parameter $\sigma$.

The prior for $\sigma$ is chosen so that the highest hazard values over the 20 year horizon (90\% quantile) are expected to be about $\rho=2$ times the lowest values (10\% quantile), but with a vague 95\% credible interval for $\rho$ of between about 1 and 16.    The function `prior_haz_sd()` uses simulation to estimate the beliefs implied by a particular Gamma prior for $\sigma$ (jointly with the prior specified for the hazard scale).

The choice of Gamma(2,5) here was arrived at through trial-and-error, until the distribution for $\rho$ (the quantity `hr` returned by `prior_haz_sd()`) matched our prior judgement.  Note that repeated simulations will give slightly different results unless the seed is set.

```{r prior_hsd}
set.seed(1)
prior_hsd <- p_gamma(2, 5)
prior_haz_sd(mspline = mspline,
             prior_hsd = prior_hsd,
             prior_hscale = prior_hscale)
```


#### Hazard ratio for treatment effects 

The prior for the treatment effect is weak, and designed only to rule out extremely implausible values, with a prior median of 1 and upper 95\% credible limit of 50 for the hazard ratio.  The function `p_hr()` translates this information to a normal prior for the log hazard ratio.  The function `prior_hr()` goes in the other direction, and summarises the prior for the hazard ratio implied by a particular normal distribution on the log scale (in the special format returned by `p_hr()`, or alternatively by the function `p_normal()`).

```{r prior_hr}
prior_loghr <- p_hr(median=1, upper=50)
prior_hr(prior_loghr)
```


#### Hazard ratio variability parameter $\tau$ in non-proportional hazards model

In a similar way, we can calibrate the Gamma prior for the parameter $\tau$, which is only used in non-proportional hazards models. This governs the size of departure from proportional hazards, i.e. the variability in the hazard ratio over time.  $\tau$ itself does not have a direct interpretation, but we can simulate hazard curves under different treatment groups, and compute the ratio $\rho$ between the "highest" (90\% quantile) and "lowest" (10\% quantile) hazard ratio over time.   If hazards are proportional, then $\rho=1$. 

Here a Gamma(2,3) prior is used, which represents a judgement that the hazard ratio could vary about 10-fold (to an order of magnitude) at different points over 20 years.

```{r prior_hr_sd}
set.seed(1)
prior_hrsd <- p_gamma(2, 3)
prior_hr_sd(mspline = mspline,                              
            prior_hsd = prior_hsd,
            prior_loghr = prior_loghr,
            prior_hscale = prior_hscale,
            prior_hrsd = prior_hrsd,
            formula = ~treat,
            nonprop = ~treat,
            newdata = data.frame(treat=1), 
            newdata0 = data.frame(treat=0))
```


# Notes on model computation

By default, `survextrap` fits models by MCMC sampling.   This can be slow, and with the default settings takes 2 minutes per run.  For model development, we can reduce the number of MCMC chains and iterations by setting `chains` and `iter` when calling `survextrap`.  These default to 4 and 2000 respectively in the `rstan` package that `survextrap` uses for computation.

An approximation to full MCMC sampling can also be specified by calling `survextrap` with `fit_method="opt"`: this runs instantly and obtains the posterior mode exactly, but crudely approximates other posterior summaries.  This is very useful for development.

When we are ready to produce the "final" results, we can switch back to MCMC with the default number of chains and iterations. 

```{r mcmc_setup}
options(mc.cores = 2) # CRAN check limits to 2 cores
# options(mc.cores = parallel::detectCores())
chains <- 4; iter <- 2000
```

The `rhat` convergence diagnostic shown when printing the results of a `survextrap` fit should be checked - this should be close to 1 if the MCMC fit has converged. 

Occasionally the fitting may report "divergent transitions".  This is a limitation of the sampling algorithm, and can happen when the posterior distribution is awkward to sample from.  See the [Stan documentation](https://mc-stan.org/misc/warnings.html) for technical details.  If there are only one or two divergent transitions, the message is probably safe to ignore, but in any case the results should be examined to ensure that they make sense.  With lots of divergent transitions, it is safer to simplify the model, e.g. by using stronger priors or fewer spline knots.


# Models for the trial data alone

## Single treatment group

First we fit two models for the control group trial data: one (`mod_con`) allowing hazard variations up to 20 years, and another (`mod_con5`) assuming a constant hazard after 5 years.   `mod_con5` is implemented by leaving out the `mspline` argument to `survextrap`, in which case the final spline knot is set to the end of the data.  All models in `survextrap` assume a constant hazard after the final spline knot.

```{r mod_con,cache=TRUE,results="hide"}
mod_con <- survextrap(Surv(years, d) ~ 1, data=control, mspline=mspline,
                      chains=chains, iter=iter,
                      prior_hscale=prior_hscale, prior_hsd = prior_hsd)
mod_con5 <- survextrap(Surv(years, d) ~ 1, data=control, 
                       chains=chains, iter=iter, 
                       prior_hscale=prior_hscale, prior_hsd = prior_hsd)
```

A plot comparing the survival and hazard curves from these models is then built.  We use the `survival()` and `hazard()` functions to extract survival and hazard estimates, and combine these to give a dataset that can be plotted with `ggplot()`. 

```{r fig_control}
surv_const <- survival(mod_con5, tmax=20) %>% mutate(model="Constant hazard")
surv_ipd <- survival(mod_con, tmax=20) %>% mutate(model="Varying hazard")
surv_single <- rbind(surv_const, surv_ipd)
haz_const <- hazard(mod_con5, tmax=20) %>% mutate(model="Constant hazard")
haz_ipd <- hazard(mod_con, tmax=20) %>% mutate(model="Varying hazard")
haz_single <- rbind(haz_const, haz_ipd)

ps <- ggplot(surv_single, aes(x=t, y=median, 
                        group=model, col=model, fill=model)) + 
    geom_ribbon(aes(ymin=lower, ymax=upper), alpha=0.2, colour=NA) +
    geom_line(lwd=1.3) + 
    geom_step(data=mod_con$km, aes(x=time, y=surv), lwd=1.3,
              inherit.aes = FALSE) +
    theme_minimal() + 
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
    xlab("Years after diagnosis") + 
    ylab("Survival probability") +
    scale_color_viridis(discrete=TRUE) + 
    scale_fill_viridis(discrete=TRUE) + 
    theme(legend.position = c(0.6, 0.8)) + labs(col=NULL, fill=NULL) +
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
    geom_vline(xintercept = max(control$years)) +
    geom_vline(xintercept = mod_con5$mspline$iknots, col="gray80", lty=2)

ph <- ggplot(haz_single, aes(x=t, y=median, 
                        group=model, col=model, fill=model)) + 
    geom_ribbon(aes(ymin=lower, ymax=upper), alpha=0.2, colour=NA) +
    geom_line(lwd=1.3) + 
    theme_minimal() + xlab("Years after diagnosis") + ylab("Hazard") +
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
    scale_color_viridis(discrete=TRUE) + 
    scale_fill_viridis(discrete=TRUE) + 
    geom_vline(xintercept = max(control$years)) +
    geom_vline(xintercept = mod_con5$mspline$iknots, col="gray80", lty=2) + 
    theme(legend.position = "none")

grid::grid.newpage()
grid::grid.draw(cbind(ggplotGrob(ps), 
                      ggplotGrob(ph)))
```

The plots show the posterior median and 95\% credible intervals. There are no data between 5 and 20 years, so the extrapolation of the trial data over this period depends on the model assumption - either that the hazard is constant, or that the hazard will change smoothly, but with an unknown direction and extent of change.   If we allow it to change, the uncertainty is appropriately greater. 


## Determining an optimal number of knots

The model seems to fit the trial data well, and the Bayesian procedure is designed to mitigate against overfitting from having too many spline knots.   Even so, we might still want reassurance that the spline function has an appropriate amount of complexity to describe the data.  Therefore we use LOOIC to compare a range of models with different numbers of spline basis functions (set via the `df` argument to `mspline_spec`).

<details>
```{r modcomp, class.source = 'fold-hide'}
dfs <- 5:12
rescomp <- as.data.frame(matrix(nrow=length(dfs), ncol=7))
for (i in seq_along(dfs)){
    msp <- mspline_spec(Surv(years, d) ~ 1, data=control, df=dfs[i])
    mod <- survextrap(Surv(years, d) ~ 1, data=control, mspline=msp,
                      chains=chains, iter=iter,
                      prior_hscale = prior_hscale, prior_hsd = prior_hsd) 
    rescompi <- cbind(list(df = dfs[i]), 
                  list(looic = mod$loo$estimates["looic","Estimate"]),
                  rmst(mod,t=5))
    rescomp[i,] <- rescompi
}
names(rescomp) <- names(rescompi)
rescomp %>% 
  mutate(looic = round(looic, 1), 
         rmf = sprintf("%s (%s, %s)", round(median,2), 
                       round(`2.5%`, 2), round(`97.5%`, 2))) %>%
  select(df, looic, rmf) %>%
  knitr::kable(col.names = c("df", "LOOIC", "Restricted mean survival (5 years)"))
```
</details>
The estimates of RMST within 5 years do not change within 0.1.   Note this is only one of many ad-hoc checks we could do - note we did not change the prior for the smoothness `prior_hsd`, which also governs the amount of smoothness of the hazard function.  LOOCV is also lower for `df=7`.  In practice we might want to switch the `df` according to what data are included, but for simplicity of illustration in this article, we stick with `df=6` for all models, which in general gives slightly stabler computation without affecting the results.

Using LOOIC for model choice may result in fitted hazard functions that wiggle a lot over time and appear to "overfit" the data - but this may not matter if we just want to use the model for estimating quantities averaged over time, such as restricted mean survival.   LOOIC is designed for _prediction_.  If instead we want to use it for _description_ of the hazard trajectory, we might want to pick a simpler model by setting `df` to a lower value. 


## Models for the trial data with a treatment effect

We now fit a joint model for both arms of the trial data, describing the treatment effect either through proportional hazards (`mod_ph`) or with `survextrap`'s novel flexible non-proportional hazards model (`mod_nph`).
```{r mod_trt,cache=TRUE,results="hide"}
mod_ph <- survextrap(Surv(years, d) ~ treat, data=cetux, 
                     mspline=mspline, 
                     chains=chains, iter=iter,
                     prior_hscale=prior_hscale, prior_hsd = prior_hsd,
                     prior_loghr=prior_loghr)
mod_nph <- survextrap(Surv(years, d) ~ treat, data=cetux, mspline=mspline, 
                      chains=chains, iter=iter,
                      nonprop = TRUE,
                      prior_hscale=prior_hscale, prior_hsd = prior_hsd,
                      prior_loghr=prior_loghr, prior_hrsd=prior_hrsd)
```

We also fit a model for the cetuximab treatment arm alone, to couple with the model `mod_con` for the control arm that we fitted earlier. 
```{r mod_cet,results="hide"}
mod_cet <- survextrap(Surv(years, d) ~ 1, data=cetuximab, mspline=mspline, 
                      chains=chains, iter=iter,
                      prior_hscale=prior_hscale, prior_hsd = prior_hsd)
```

The survival curves for these three models are now compared.

<details>
```{r fig_trt, class.source = 'fold-hide'}
surv_ph <- survival(mod_ph, tmax=7) %>% mutate(Model="Proportional hazards")
surv_nph <- survival(mod_nph, tmax=7) %>% mutate(Model="Non-proportional hazards")
surv_cet <- survival(mod_cet, tmax=7) %>% 
    mutate(Model="Separate fits", treat="Cetuximab") %>% relocate(treat)
surv_con <- survival(mod_con, tmax=7) %>% 
    mutate(Model="Separate fits", treat="Control") %>% relocate(treat)
surv_trt <- rbind(surv_ph, surv_nph, surv_cet, surv_con)

ggplot(surv_trt, aes(x=t, y=median, col=Model, lty=treat)) + 
    geom_step(data=mod_ph$km, aes(x=time, y=surv, lty=treat), lwd=1.3,
              inherit.aes = FALSE) +
    geom_line(lwd=1.1) + 
    labs(lty="Treatment", col=NULL) +
    theme_minimal() +
    theme(legend.position = c(0.6, 0.8)) +
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
    xlab("Years after diagnosis") + 
    ylab("Survival probability") +
    scale_color_viridis(discrete=TRUE) + 
    geom_vline(xintercept = max(cetux$years)) +
    geom_vline(xintercept = mod_con5$mspline$iknots, col="gray80", lty=2)
```
</details>


The results and goodness-of fit are compared between these three models.  We compare estimates of restricted mean survival for the control group (`rmst()`), and the increase in mean survival time for the treated group (`irmst()` ). Both of these are restricted to the 5-year trial horizon (`t=5`).

Note that when using functions like `rmst()` or `survival()` which extract outputs from models with covariates, we supply `newdata` to say what covariate values we want the outputs to be computed for.  In the call to `rmst()` we told it to compute for the control group.  We left out `newdata` in the call to `irmst()` because, for models with one factor covariate with two levels, it will calculate, by default, the mean for the second level minus the mean for the first level.

<details>
```{r table_trt, class.source = 'fold-hide'}
comp_fn <- function(mod){
    rm <- rmst(mod, niter=1000, newdata=data.frame(treat="Control"), t=c(5,20)) %>% 
        rename(rm_med="median",rm_lower="2.5%", rm_upper="97.5%") %>% 
        select(t, rm_med, rm_lower, rm_upper)
    irm <- irmst(mod, niter=1000, t=c(5,20)) %>%
        rename(ir_med="median",ir_lower="2.5%", ir_upper="97.5%") %>%
        select(ir_med, ir_lower, ir_upper)
    cbind(name=deparse(substitute(mod)), rm, irm, 
          list(looic = mod$loo$estimates["looic","Estimate"]))
}

rmst_cet <- rmst(mod_cet, t=c(5,20))
rmst_con <- rmst(mod_con, t=c(5,20)) %>%
  rename(rm_med="median",rm_lower="2.5%", rm_upper="97.5%") %>% 
  select(t, rm_med, rm_lower, rm_upper)
irmst_sep <- attr(rmst_cet,"sample")[[1]] - attr(rmst_con,"sample")[[1]]
irmst_sep <- posterior::summarise_draws(irmst_sep, median,
                                        ~quantile(.x, probs=c(0.025, 0.975))) %>%
  rename(t=variable, ir_med="median",ir_lower="2.5%", ir_upper="97.5%") %>% 
  select(ir_med, ir_lower, ir_upper)
loo_sep <- mod_con$loo$estimates["looic","Estimate"] +
  mod_cet$loo$estimates["looic","Estimate"]
comp_sep <- cbind(list(name="mod_sep"), rmst_con, irmst_sep, list(looic=loo_sep))

trtcomp <- rbind(comp_fn(mod_ph), comp_fn(mod_nph), comp_sep)
trtcompf <- trtcomp %>%
  arrange(t) %>%
  mutate(rm = sprintf("%s (%s,%s)", round(rm_med,2),
                      round(rm_lower,2), round(rm_upper,2)),
         ir = sprintf("%s (%s,%s)", round(ir_med,2),
                      round(ir_lower,2), round(ir_upper,2)),
         looic = round(looic),
         model = forcats::fct_recode(name, 
                                     "(2a) Proportional hazards"="mod_ph",
                                     "(2b) Non-proportional hazards"="mod_nph",
                                     "(2c) Separate arms"="mod_sep"
         )) %>%
  select(model, t, rm, ir, looic) %>%
  tibble::remove_rownames()
knitr::kable(trtcompf, col.names=c("Model","Time horizon",
                                   "Restricted mean survival",
                                   "Increase in restricted mean survival",
                                   "LOOIC"))
```
</details>

There is very little difference between the fit of the three models. 

The following code formats the table in LaTeX format (for the paper).
```{r latex_table_trt}
write.table(trtcompf, sep="  &  ", eol="\\\\\n", quote=FALSE, row.names = FALSE)
```


# External data: registry data on control group

In this example, there are registry data giving counts of $r_j$ survivors to the next year out of $n_j$ alive at year $j$, from from $j=5$ years onwards.  This is supplied in the `survextrap` package as `cetux_seer`. 
```{r cetux_seer}
head(cetux_seer)
```
This data can be illustrated as annual hazard estimates, calculated as $-\log(r_j/n_j)$, shown as the thin step function in the plot below. 

We fit a model which assumes that people in the registry data have the same survival as the control group in the trial.  To allow for potential variations in the hazard shape between each of the years from 6 to 25 for which registry data are reported, we add extra knots at 10, 15 and 20 years.   

> **Note**: if instead there had been only one row in the external data, giving survivors to 25 years from those alive at 6 years, it would be impossible to identify variations in the hazard _within_ the period from 6 to 25 years, so using multiple knots inside this period would be futile.

```{r mod_seer,cache=TRUE,results="hide"}
mspline_seer <- mspline_spec(Surv(years, d) ~ 1, data=cetux, df=6, 
                             add_knots=c(10,15,20))
mod_seer <- survextrap(Surv(years, d) ~ 1, data=control, 
                       external=cetux_seer, mspline=mspline_seer, 
                       chains=chains, iter=iter, 
                       prior_hscale=prior_hscale, prior_hsd = prior_hsd)
```

The results are illustrated, and compared to the model `mod_con` for the trial data alone that allowed the hazard to change after the trial.  The registry data makes the hazard extrapolations much more confident. 

<details>
```{r fig_registry, class.source = 'fold-hide'}
haz_con <- hazard(mod_con) %>%
    mutate(Model="No external data")
haz_seer <- hazard(mod_seer,tmax=30) %>% 
    mutate(Model="With registry data")

haz_plot_reg <- rbind(haz_con, haz_seer) %>% filter(t < 26)

ggplot(haz_plot_reg, aes(x=t, y=median, col=Model, fill=Model)) + 
    geom_ribbon(aes(ymin=lower, ymax=upper), alpha=0.2, colour=NA) +
    geom_line() +
    geom_step(data=cetux_seer, aes(x=start, y=haz), col="gray30", inherit.aes = FALSE) +
    geom_step(data=cetux_seer, aes(x=start, y=haz_lower), col="gray70", lty=2, inherit.aes = FALSE) +
    geom_step(data=cetux_seer, aes(x=start, y=haz_upper), col="gray70", lty=2, inherit.aes = FALSE) + 
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
    xlab("Years after diagnosis") + 
    ylab("Hazard") +
    scale_color_viridis(discrete=TRUE) + 
    scale_fill_viridis(discrete=TRUE) + 
    geom_vline(xintercept = max(control$years))  + 
    theme(legend.position = c(0.4, 0.9), 
          legend.background=element_blank()) + 
    labs(col=NULL, fill=NULL)
```
</details>


## Knot selection for external data 

The hazard seems to fit the external data, even with just one knot designed to capture hazard trajectory changes after the trial, placed at 20 years.  For some reassurance, we examine some alternative knot specifications, to ensure that we have captured all variations that we need in order to estimate survival up to 20 years. 

Placing knots at 10,15 and 20 years gives the lowest (best) LOOIC from these alternatives.  Adding more knots is judged to be over-fitting, which hampers predictive ability, and fewer knots do not represent all hazard variations.  The changes in the RMST estimates between these models are small though. 

<details>
```{r modcomp_seer, cache=TRUE}
knot_list <- list(c(10,20), 
                  c(20),
                  seq(8,20,by=2))

comp_fn <- function(mod){
  rm <- rmst(mod, niter=1000, newdata=data.frame(treat="Control"), t=c(20)) %>% 
    rename(rm_med="median",rm_lower="2.5%", rm_upper="97.5%") %>% 
    mutate(rmf = sprintf("%s (%s, %s)", round(rm_med,2), 
                         round(rm_lower,2), round(rm_upper,2))) %>%
    select(t, rmf)
  cbind(name=deparse(substitute(mod)), rm,  
        list(looic = round(mod$loo_external$estimates["looic","Estimate"])))
}

res <- vector(3, mode="list")
for (i in 1:3){
  mspex <- mspline_spec(Surv(years, d) ~ treat, data=cetux, df=6, 
                         add_knots = knot_list[[i]])
  mod <- survextrap(Surv(years, d) ~ 1, data=control, 
                    external=cetux_seer, mspline=mspex, 
                    chains=chains, iter=iter, 
                    prior_hscale=prior_hscale, prior_hsd = prior_hsd)
  res[[i]] <- comp_fn(mod)
}

do.call("rbind", c(list(comp_fn(mod_seer)), res)) %>%
  mutate(knots = c("10,15,20",
                   unlist(lapply(knot_list, paste, collapse=",")))) %>%
  select(knots, rmf, looic) %>%
  knitr::kable(col.names=c("Additional knots",
                           "Restricted mean survival over 20 years","LOOIC"))
```
</details>




# Population data in an additive hazards model

An alternative way of including external data is through additive hazards (or relative survival), using the `backhaz` argument to supply a data frame that describes the background hazard at all times.  This describes the population of the USA but matched to the trial patients by age, sex and calendar year, and is supplied in the `survextrap` package as `cetux_bh`.

```{r cetux_bh}
head(cetux_bh)
```
The remaining data (trial and registry) are used to infer the excess hazard for head and neck cancer patients.  A model is fitted which assumes the overall hazard for the trial and registry patients is the sum of the background hazard (fixed from the data in `backhaz`) and an unknown excess hazard to be estimated.

```{r mod_seer_pop, results="hide"}
mod_seer_pop <- survextrap(Surv(years, d) ~ 1, data=control, mspline = mspline_seer,
                           external=cetux_seer, backhaz=cetux_bh, 
                           chains=chains,iter=iter,
                           prior_hscale=prior_hscale, prior_hsd = prior_hsd)
```

<details>
```{r fig_registry_pop20, class.source = 'fold-hide'}
haz_seer <- hazard(mod_seer,tmax=20) %>% mutate(Model="Without population data")
haz_seer_pop <- hazard(mod_seer_pop,tmax=20) %>% mutate(Model="With population data")
haz_plot_reg_pop20 <- rbind(haz_seer, haz_seer_pop)

ggplot(haz_plot_reg_pop20, aes(x=t, y=median, col=Model, fill=Model)) + 
    geom_ribbon(aes(ymin=lower, ymax=upper), alpha=0.2, colour=NA) +
    geom_line() +
    geom_step(data=cetux_seer  %>% filter(stop < 20), 
              aes(x=start, y=haz), inherit.aes = FALSE) +
    geom_step(data=cetux_seer  %>% filter(stop < 20), 
              aes(x=start, y=haz_lower), col="gray60", lty=2, inherit.aes = FALSE) +
    geom_step(data=cetux_seer  %>% filter(stop < 20), 
              aes(x=start, y=haz_upper), col="gray60", lty=2, inherit.aes = FALSE) + 
    geom_step(data=cetux_bh %>% filter(time < 20), 
              aes(x=time, y=hazard), inherit.aes = FALSE, colour="darkgreen") +
    annotate(geom="text", x=10, y=0, label="Population mortality", colour="darkgreen", size=2) +
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
    xlab("Years after diagnosis") + 
    ylab("Hazard") +
    scale_color_viridis(discrete=TRUE) + 
    geom_vline(xintercept = max(control$years)) +
    theme(legend.position = c(0.8, 0.9), 
          legend.background=element_blank()) + 
    labs(col=NULL, fill=NULL)
```
</details>

Including the population data does not make much difference to the estimated hazards over 20 years after diagnosis. 

The benefit of the population data in this example is to inform extrapolations over a longer time horizon than the 25 year horizon of the registry data.   We now consider extrapolations up to 40 years.   To allow the hazard shape to change up to 40 years, we extend the spline specification to include two further knots at 30 and 40 years, in addition to the ones already placed at 10, 15 and 20 years. 

Two models are then fitted, to show the impact of the population data. The first includes the trial and registry data.

```{r mod_seer40, results="hide"}
mspline40 <- mspline_spec(Surv(years, d) ~ treat, data=cetux, df=6, 
                          add_knots=c(10, 15, 20, 30, 40))
mod_seer40 <- survextrap(Surv(years, d) ~ 1, data=control, mspline = mspline40,
                         external=cetux_seer, 
                         chains=chains, iter=iter, 
                         prior_hscale=prior_hscale, prior_hsd = prior_hsd)
```

The second includes the trial, registry and population data.
<a id="popplot"></a>
```{r mod_registry_pop40, results="hide"}
mod_seer_pop40 <- survextrap(Surv(years, d) ~ 1, data=control, mspline = mspline40,
                           external=cetux_seer, backhaz=cetux_bh,
                           chains=chains, iter=iter, 
                           prior_hscale=prior_hscale, prior_hsd = prior_hsd)
```

The population data "anchors" the extrapolation - the hazard is constrained to be no less than that of the population, and the resulting estimates become more precise.  There is still a lot of uncertainty, since (through the knots at 30 and 40 years) we have allowed for the possibility that the excess hazard changes beyond the point (25-26 years) where we actually have data about it. 

<details>
```{r fig_registry_pop40, class.source = 'fold-hide'}
haz_seer40 <- hazard(mod_seer40,tmax=40) %>% 
    mutate(Model="With registry data")
haz_seer_pop40 <- hazard(mod_seer_pop40,tmax=40) %>% 
    mutate(Model="With registry and population data")
haz_plot_reg_pop <- rbind(haz_seer40, haz_seer_pop40)
ggplot(haz_plot_reg_pop, aes(x=t, y=median, col=Model, fill=Model)) + 
    ylim(0,0.5) +
    geom_ribbon(aes(ymin=lower, ymax=upper), alpha=0.2, colour=NA) +
    geom_line() +
    geom_step(data=cetux_seer, aes(x=start, y=haz), inherit.aes = FALSE) +
    geom_step(data=cetux_seer, aes(x=start, y=haz_lower), col="gray60", lty=2,
              inherit.aes = FALSE) +
    geom_step(data=cetux_seer, aes(x=start, y=haz_upper), col="gray60", lty=2,
              inherit.aes = FALSE) + 
    geom_step(data=cetux_bh %>% filter(time < 40), 
              aes(x=time, y=hazard), inherit.aes = FALSE, colour="darkgreen") +
    annotate(geom="text", x=33, y=0.1, label="Population mortality", colour="darkgreen", size=2) +
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
    xlab("Years after diagnosis") + 
    ylab("Hazard") +
    scale_color_viridis(discrete=TRUE) + 
    theme(legend.position = c(0.4, 0.9), 
          legend.background=element_blank()) + 
    labs(col=NULL, fill=NULL) +
    geom_vline(xintercept = max(control$years)) 
```
</details>


# Mixture cure model

The extrapolation of the excess hazard is still fairly uncertain, even when both registry and population data are included, since the data informing the excess hazard finish at 25 years.

In many clinical applications, it is thought that the excess hazard for a population with a disease will decrease over time, particularly as people grow older and their all-cause mortality increases.    We examine a couple of ways to use this kind of judgement to improve survival extrapolations using `survextrap`.   The specific assumptions used here may or may not be realistic for these patients: this case study is intended as a demonstration of methods, rather than a full, clinically-informed investigation of head and neck cancer survival.

One way of modelling a decreasing hazard is through a _mixture cure_ model.   This can also be used to model a decreasing _excess_ hazard in an additive hazards model.  This is simply specified by setting `cure=TRUE` in the call to `survextrap`.

```{r mod_cure, results="hide"}
mod_cure <- survextrap(Surv(years, d) ~ 1, data=control, mspline = mspline40,
                       external=cetux_seer, backhaz=cetux_bh, 
                       cure=TRUE, 
                       chains=chains, iter=iter,
                       prior_hscale=prior_hscale, prior_hsd = prior_hsd)
```

Optionally, a Beta prior could also be specified for the "cure fraction" through e.g. `survextrap(..., prior cure = p_beta(1,10))`, though the default uniform prior is used here.  The cure fraction is the proportion of people who will never experience the event of interest.  The survival curve decreases to an asymptote at the cure fraction, hence (informally) the cure fraction can sometimes be inferred from data by examining the survival curve.  Often however it is weakly informed by data.

In this example, the excess hazard for head and neck cancer patients is still substantial by the end of the registry data (25 years), so there is little evidence of "cure" from the data.   Therefore the extrapolations from the fitted cure model are very uncertain [(see below)](#cureplot), and largely overlap with those from the equivalent model for the same data where cure was not assumed [(see above)](#popplot).  In addition, the cure model makes a strong parametric assumption for the shape of how the hazard decreases over time.


# Elicited data to inform long-term extrapolations

An alternative way to build in judgements about longer-term survival is through expert elicitation, as discussed in the [priors vignette](priors.html). 

Here, suppose we are confident that by 40 years, the survival of head and neck cancer patients will be similar to that of the general population.  Instead of using a parametric cure model, we can create an artificial external dataset that contains the same information as our judgement.  Here the annual population survival probability is computed from the hazard stored in `cetux_bh`, and stored in `pops`.  

```{r elic}
cetux_bh <- cetux_bh %>% 
  mutate(surv1 = exp(-hazard*1))
pops <- cetux_bh$surv1[39]
```

A `Beta(n*pops, n(1-pops))` distribution for any `n` will have a mean of `pops`, here `r round(pops,2)`.  We then set `n` to control the amount of uncertainty around this prior guess.

```{r elic_n}
qbeta(c(0.025, 0.5, 0.975), 1000*pops, 1000*(1-pops))
```

The information in this Beta prior is equivalent to having observed `1000*pops` survivors out of 1000 people.  This gives a numerator and denominator `n` and `r` for survival between `start=39` and `stop=40` years.  We use these to construct a dataset in the format needed by the `external` argument to `survextrap`.

```{r data_elic}
data_elic <- data.frame(
  start=39, stop=40,
  n = 1000, 
  r = round(1000*pops),
  treat = "Control"
)
```

This can then be joined to the registry data `cetux_seer`, to give an external dataset to supply to `survextrap`. 

```{r mod_cure_elic, results="hide"}
pop_cured <- 
  cetux_seer %>% full_join(data_elic, by = join_by(start, stop, r, n, treat))
mod_cure_elic <- survextrap(Surv(years, d) ~ 1, data=control, mspline=mspline40, 
                            external=pop_cured, backhaz=cetux_bh, 
                            chains=chains, iter=iter,
                            prior_hscale=prior_hscale, prior_hsd = prior_hsd)
```

The extrapolations of excess hazard from this model are clearly informed by the elicited survival probability, and pulled closer to the population data.  If we had expressed more confidence in the prior guess (through a bigger `n`) the posterior estimates of the hazard function from the model would have been pulled even closer to the population data.
<a id="cureplot"></a>
<details>
```{r fig_cure, class.source = 'fold-hide'}
haz_cure <- hazard(mod_cure, tmax=40) %>% mutate(Model="Mixture cure model")
haz_cure_elic <- hazard(mod_cure_elic, tmax=40) %>% mutate(Model="Elicited survival")
haz_plot_cure <- rbind(haz_cure, haz_cure_elic)

ggplot(haz_plot_cure, aes(x=t, y=median, col=Model, fill=Model)) + 
    ylim(0,0.5) +
    geom_ribbon(aes(ymin=lower, ymax=upper), alpha=0.2, colour=NA) +
    geom_line() +
    geom_step(data=cetux_seer, aes(x=start, y=haz), inherit.aes = FALSE) +
    geom_step(data=cetux_seer, aes(x=start, y=haz_lower), col="gray60", lty=2,
              inherit.aes = FALSE) +
    geom_step(data=cetux_seer, aes(x=start, y=haz_upper), col="gray60", lty=2,
              inherit.aes = FALSE) + 
    geom_step(data=cetux_bh %>% filter(time < 40), 
              aes(x=time, y=hazard), inherit.aes = FALSE, col="darkgreen") +
    annotate(geom="text", x=20, y=0, label="Population mortality", col="darkgreen", size=2) +
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
    xlab("Years after diagnosis") + 
    ylab("Hazard") +
    scale_color_viridis(discrete=TRUE) + 
    geom_vline(xintercept = max(control$years)) +
    theme(legend.position = c(0.4, 0.9), 
          legend.background=element_blank()) + 
    labs(col=NULL, fill=NULL) 
```
</details>

# Table of mean survival times

A table of results for the paper is now constructed and formatted.   We compare restricted mean survival times, over various time horizons, estimated from the various models fitted to the control group data. 

The models are categorised by what data are included, and what assumptions are made about survival beyond the data.

<details>
```{r rmst_con, cache=TRUE, class.source = 'fold-hide'}
ni <- 4000

rcon <- rmst(mod_con, t=5, niter=ni) %>% 
  mutate(data="Trial",extrap="No extrapolation")
rcon20 <- rmst(mod_con, t=20, niter=ni) %>% 
  mutate(data="Trial", extrap="Uncertain hazard")
rcon520 <- rmst(mod_con5, t=20, niter=ni) %>% 
  mutate(data="Trial", extrap="Constant hazard")
rreg20 <- rmst(mod_seer, t=20, niter=ni) %>% 
  mutate(data="Trial,registry",extrap="No extrapolation")
rreg40 <- rmst(mod_seer40, t=40, niter=ni)  %>% 
  mutate(data="Trial,registry", extrap="Uncertain hazard")
rpop40 <- rmst(mod_seer_pop40, t=40, niter=ni) %>%
  mutate(data="Trial,registry,population",extrap="Uncertain excess hazard")
rcure40 <- rmst(mod_cure, t=40, niter=ni) %>% 
  mutate(data="Trial,registry,population", 
         extrap="Mixture cure")
relic40 <- rmst(mod_cure_elic, t=40, niter=ni) %>%
  mutate(data="Trial,registry,population", 
         extrap="Elicited survival")

modlabs <- c("(1a)", "(1a)", "(1b)", "(1c)", "(1d)","(1e)", "(1f)", "(1g)")

rs <- list(rcon, rcon20, rcon520, rreg20, rreg40, rpop40, rcure40, relic40) %>%
  purrr::reduce(full_join, by = join_by(variable, t, median, `2.5%`, `97.5%`, data, extrap)) %>%
  mutate(modlabs=modlabs,
         rmst = sprintf("%s (%s, %s)", round(median,2), 
                        round(`2.5%`,2), round(`97.5%`,2))) %>%
  select(modlabs, data, extrap, t, rmst)
knitr::kable(rs, col.names=c("Model", "Observed data","Extrapolation assumptions",
                             "Time horizon","Restricted mean survival"))
```
</details>

The following code formats the table in LaTeX format (for the paper).
```{r latex_table_rs}
write.table(rs, sep="  &  ", eol="\\\\\n", quote=FALSE, row.names = FALSE)
```


# Waning models for long-term treatment effects

Finally, we estimate the effect of treatment with cetuximab on long-term survival.  We extend two of the models with external data (fitted earlier to just the control group) to include a treatment effect and fit them to the full trial data `cetux`.  Firstly, the model with trial and registry data, and secondly, the model with all three data sources (trial, registry and population, but no "cure" assumption).  The proportional hazards model is assumed, which was adequate to describe the short-term data.

```{r mod_full, cache=TRUE, results="hide"}
mod_full_reg <- survextrap(Surv(years, d) ~ treat, data=cetux, mspline = mspline40,
                           external=cetux_seer, 
                           chains=chains, iter=iter, 
                           prior_hscale=prior_hscale, prior_hsd = prior_hsd, 
                           prior_loghr = prior_loghr)
mod_full_pop <- survextrap(Surv(years, d) ~ treat, data=cetux, mspline = mspline40,
                           external=cetux_seer, backhaz=cetux_bh,
                           chains=chains, iter=iter, 
                           prior_hscale=prior_hscale, prior_hsd = prior_hsd, 
                           prior_loghr = prior_loghr)
```

The restricted mean survival for the control group, and the increase in the restricted mean survival given by cetuximab, is estimated under various "treatment effect waning" assumptions.   Essentially these assume that the log hazard ratio for a "treated" group decreases linearly over some time period from the "treated" estimate to the "control" estimate.   This is done by mixing the posteriors of treated and control estimates obtained from the same model, in this case `mod_full_pop` (see the [methods vignette](methods.html#waning)).

The calls to `rmst` and `irmst` are modified to include a `wane_period` (vector defining the start and end of the waning period).  We also supply `newdata` and `newdata0`, which indicate respectively what covariate values in the model correspond to "treated" and "control" groups.

```{r waning, cache = TRUE}
nd <- data.frame(treat=c("Cetuximab"))
nd0 <- data.frame(treat=c("Control"))
ind <- data.frame(treat=c("Control","Cetuximab"))
ind0 <- data.frame(treat=c("Control","Control"))
tmax <- max(cetux$years)

ni <- 1000

## 20 year absolute survival
rmnwr <- rmst(mod_full_reg, t=20, niter=ni) %>% mutate(wane=0)
rmnw <- rmst(mod_full_pop, t=20, niter=ni) %>% mutate(wane=0)
rmw6 <- rmst(mod_full_pop, t=20, niter=ni, newdata=nd, newdata0=nd0, 
             wane_period = c(tmax, 6)) %>% mutate(wane=6)
rmw20 <- rmst(mod_full_pop, t=20, niter=ni, newdata=nd, newdata0=nd0, 
              wane_period = c(tmax, 20)) %>% mutate(wane=20)

## 20 year incremental survival for cetuximab
irmnwr <- irmst(mod_full_reg, t=20, niter=ni) %>% mutate(wane=0)
irmnw <- irmst(mod_full_pop, t=20, niter=ni) %>% mutate(wane=0)
irmw6 <- irmst(mod_full_pop, t=20, niter=ni, newdata = ind, newdata0 = ind0, 
               wane_period = c(tmax,6), wane_nt = 5) %>% mutate(wane=6)
irmw20 <- irmst(mod_full_pop, t=20, niter=ni, newdata = ind, newdata0 = ind0, 
                wane_period = c(tmax,20), wane_nt = 5) %>% mutate(wane=20)

rmd <- rbind(rmnwr, rmnw, rmw6, rmw20) %>% 
    filter(treat=="Control") %>%
    rename(rm="median",rmlower="2.5%",rmupper="97.5%") %>%
    mutate(data=c("reg","pop"),
           rmf = sprintf("%s (%s, %s)", round(rm,2), 
                         round(rmlower,2), round(rmupper,2))) %>%
    select(data, wane, rmf) 
irmd <- rbind(irmnwr, irmnw, irmw6, irmw20) %>% 
    rename(irm="median",irmlower="2.5%",irmupper="97.5%") %>%
    mutate(data=c("reg","pop","pop","pop"),
           irmf = sprintf("%s (%s, %s)", round(irm,2), 
                          round(irmlower,2), round(irmupper,2))) %>%
    select(data, wane, irmf)
res_wane <- rmd %>% 
    select(-wane) %>%
    right_join(irmd, by=c("data")) %>%
  mutate(model=c("(2d) No waning","(2e) No waning",
                 "(2e) 5 to 6 years", "(2e) 5 to 20 years"),
           looic = "",
           rmf = tidyr::replace_na(rmf, "")) %>%
    select(model, rmf, irmf, looic)
knitr::kable(res_wane %>%
             select(-looic) %>%
             mutate(data = c("Trial,registry",rep("Trial,registry,population",3))) %>%
             relocate(data),
             col.names=c("Data","Model","Restricted mean survival",
                         "Increase in restricted mean survival"))
```
Table in LaTeX format (for the paper).
```{r latex_wane}
write.table(res_wane, eol="\\\\\n", sep="  &  ", row.names = FALSE, quote=FALSE)
```


Using all three data sources, with no waning, the incremental restricted mean survival over 20 years is estimated as `r irmd$irmf[2]`.  This reduces to `r irmd$irmf[4]` when waning is applied gradually from 6 to 20 years, and even further to `r irmd$irmf[3]` when the effect is assumed to wane rapidly from 5 to 6 years.
